---
title: "Lalonde exploratory"
output: html_notebook
---

Exploratory analysis of the Lalonde / Dehejia data set.

```{r}
#devtools::install_github("jjchern/lalonde")
library(lalonde)
library(ggplot2)
library(tidyverse)


```

# Lalonde RCT: male participants

The training program ran in 1976-1977. 
earnings 1978 is the outcome measure.
Ppl that enrolled before jan 1976, or were still in the program in jan 1978 were excluded.
treat is treatment, rest is pre-treatment.

```{r}
# lalonde sample (RCT, male participants, no re74)
#nsw

table(nsw$treat)
```
## unadjusted score

```{r}
lmfit <- lm(re78 ~ treat, data = nsw)

summary(lmfit)
```

## adjusted score

```{r}
lmfit <- lm(re78 ~ treat +  age + I(age^2) + education + black +
              hispanic + married + nodegree +  re75, data = df)

summary(lmfit)
```
We estimate treatment effect at +1636 dollar.
Uncertainty is large! Low power.

```{r}
ggplot(nsw, aes(x= education, y = re78, group = factor(treat), col = factor(treat))) +
  geom_point() + geom_smooth()
```

# Dehejia-Wahba RCT + RE74

Added 1974 earnings, subset on obs with this var.

## unadjusted score


```{r}
# Dehejia-Wahba Sample (male participants, with re74 --> reduces #obs)
table(nsw_dw$treat)
```

```{r}
lmfit <- lm(re78 ~ treat, data = nsw_dw)

summary(lmfit)
```
The so-called "Benchmark unbiased treatment effect" is $1794.

Dehejia & Wahba show that it is possible to get close to this number using observational data & propensity scores.

## adjusted

```{r}
lmfit <- lm(re78 ~ treat +  age + I(age^2) + education + black +
              hispanic + married + nodegree +  re74 + re75, data = nsw_dw)

summary(lmfit)
```

Adjusted treatment effect: +1676 dollar.
Apparently the adjusters do not add precision.

# Propensity scores

For the constructed dataset, we calculate the propensity score.

```{r}
df_constr <- lalonde::nsw_dw %>% 
    filter(treat == 1) %>% 
    bind_rows(lalonde::cps_controls)
```

```{r}
ps.fit.1 <- glm (treat ~ age + I(age^2) + education + black + hispanic + married + nodegree, data=df_constr, family=binomial(link="logit"))

summary(ps.fit.1)
```
```{r}
df_constr$pscores <- predict (ps.fit.1, type="response")
```

Probability of treatment is super low if we add a huge number of control cases.
So whats up with that?
We can check the model only on the treated cases?
Check what threshold we should pick to predict "treated" for these cases?


```{r}
df_constr$treat_pred <- ifelse(df_constr$pscores >= 0.1, 1, 0)

table(df_constr[df_constr$data_id == "Dehejia-Wahba Sample",]$treat_pred, 
      df_constr[df_constr$data_id == "Dehejia-Wahba Sample",]$treat)
```



```{r}
ggplot(df_constr, aes(x= age, y = pscores, group = black, col = black)) + geom_point() + geom_smooth()
```

```{r}
df_constr %>% filter(pscores > 0.3) %>% summary()

#df_constr %>% filter(pscores < 0.05) %>% summary()
```

```{r}
df_constr <- df_constr %>% mutate(has_re74 = ifelse(re74 > 0, 1, 0)) %>%
  mutate(has_re75 = ifelse(re75 > 0, 1, 0)) %>% 
  mutate(has_re78 = ifelse(re78 > 0, 1, 0))

```

```{r}
table(df_constr$has_re74, df_constr$has_re75)
table(df_constr$has_re74, df_constr$has_re78)
table(df_constr$has_re75, df_constr$has_re78)
```

